---
title: "Untitled"
author: "David Verry"
date: "29 August 2021"
output: html_document
---

## READ IN LIBRARIES

```{r}
library(tidyverse)
library(gganimate)
library(ggrepel)
```


## READ IN DATA, COUNT AND ORDER ALL CHASES BETWEEN ALL TEAMS IN PL HISTORY

```{r}
# This is where the data is from: http://www.football-data.co.uk/englandm.php

# Create blank summary dataframe to fill with the loop
all_gd_pts_evolution_all_seasons = data.frame()
all_gd_chases_all_seasons = data.frame()
all_pts_chases_all_seasons = data.frame()
all_gd_and_pts_chases_all_seasons = data.frame()
biggest_gd_chases_by_season = data.frame()
first_year = 1994

## LOTS OF MANUAL CHANGES NEEDED TO MOVE FROM GAMEWEEK TO DATE BASED CALCS
#  AND VICE VERSA


# Loop to calculate running goal difference and points
# And then caclulate chases
while (first_year < 2021) {
  
  # Define year and create URL for that year
  first_year = first_year+1
  second_year = first_year+1
  first_year_trimmed = str_sub(as.character(first_year), 3, 4)
  second_year_trimmed = str_sub(as.character(second_year), 3, 4)
  url = paste("http://www.football-data.co.uk/mmz4281/", 
              first_year_trimmed, second_year_trimmed, "/E0.csv", sep = "")

# Create running points and goal diff totals by team
all_gd_pts_evolution = read_csv(url) %>%
    rename(date = Date) %>%
    filter(date != "") %>%
    select(1:10) %>%
    pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = "venue", values_to = "Team") %>%
    mutate(gd = case_when(venue == "HomeTeam" ~ (FTHG - FTAG),
                                       venue == "AwayTeam" ~ (FTAG - FTHG)),
           pts = case_when(venue == "HomeTeam" & FTHG > FTAG ~ 3,
                              venue == "HomeTeam" & FTHG < FTAG ~ 0,
                              venue == "AwayTeam" & FTAG > FTHG ~ 3,
                              venue == "AwayTeam" & FTAG < FTHG ~ 0,
                              TRUE ~ 1),
           Team = paste(Team, first_year_trimmed, second_year_trimmed,
                        sep = "_")) %>%
    group_by(Team) %>%
    mutate(Gameweek = row_number(),
           gd_running = cumsum(gd),
           pts_running = cumsum(pts)) %>%
    select(date, Gameweek, Team, gd_running, pts_running) %>%

    # For chases by date, rather than gameweek, every club needs a total for...
    # every game date throughout the season, based on games they have played...
    # to that point. This creates full time series of dates for every club...
    # thne fills everything downwards
    ungroup() %>%
    mutate(date = lubridate::dmy(date)) %>%
    complete(date, nesting(Team)) %>%
    group_by(Team) %>%
    fill(everything(), .direction = c("down")) %>%
    replace(is.na(.), 0)


# Calculate difference between every team at every timestep
all_gd_pts_deficits = all_gd_pts_evolution %>%
  # Count your chases by either Gameweek...
    #select(-date) %>%
    #distinct() %>%
    #left_join(x = ., y = ., by = "Gameweek") %>%
  # or Date
    select(-Gameweek) %>%
    distinct() %>%
    left_join(x = ., y = ., by = "date") %>%
  
    filter(Team.x != Team.y) %>%
    mutate(first_team_deficit = gd_running.y - gd_running.x,
           first_team_deficit_pts = pts_running.y - pts_running.x) %>%
    left_join(x = ., y = ., by = c("Team.x" = "Team.x", "Team.y" = "Team.y")) %>%
    rename(first_team = Team.x, second_team = Team.y, 
           first_team_gd_1 = gd_running.x.x, second_team_gd_1 = gd_running.y.x,
           first_team_gd_2 = gd_running.x.y, second_team_gd_2 = gd_running.y.y,
           first_team_deficit_1 = first_team_deficit.x, first_team_deficit_2 = first_team_deficit.y,
           
           first_team_pts_1 = pts_running.x.x, second_team_pts_1 = pts_running.y.x,
           first_team_pts_2 = pts_running.x.y, second_team_pts_2 = pts_running.y.y,
           first_team_deficit_pts_1 = first_team_deficit_pts.x, 
           first_team_deficit_pts_2 = first_team_deficit_pts.y,
           
           # Comment one of these out as required
           #gameweek_1 = Gameweek.x, gameweek_2 = Gameweek.y,
           date_1 = date.x, date_2 = date.y
           ) %>%
  
    select(first_team, second_team, 
           #Comment one of these out as required
           #gameweek_1, 
           date_1,
           
           first_team_gd_1, second_team_gd_1, first_team_deficit_1,
           first_team_pts_1, second_team_pts_1, first_team_deficit_pts_1,
           #Comment one of these out as required
           #gameweek_2, 
           date_2,
           
           first_team_gd_2, second_team_gd_2, first_team_deficit_2,
           first_team_pts_2, second_team_pts_2, first_team_deficit_pts_2) %>%
    # Comment one of these out as required
    #filter(gameweek_2 > gameweek_1) %>%
    filter(date_2 > date_1) %>%
      
    mutate(swing = first_team_deficit_1 - first_team_deficit_2,
           swing_pts = first_team_deficit_pts_1 - first_team_deficit_pts_2,
           chase = first_team_deficit_1,
           chase_pts = first_team_deficit_pts_1,
           # Comment one of these out as required
           #total_gameweeks = gameweek_2 - gameweek_1
           total_days = date_2 - date_1
           )


    
    
# filter so you look at genuine goal diff chases i.e. one team overtakes another
# I.e. there is an inital (positive) gap,
# the second gap is smaller than the first one
# and the second gap is less than/equal to zero
all_gd_chases = all_gd_pts_deficits %>%
    filter(first_team_deficit_1 >0,
           first_team_deficit_2 < first_team_deficit_1,
           first_team_deficit_2 <= 0)

# filter so you look at genuine points chases i.e. one team overtakes another
# I.e. there is an inital (positive) gap,
# the second gap is smaller than the first one
# and the second gap is less than/equal to zero
all_pts_chases = all_gd_pts_deficits%>%
    filter(first_team_deficit_pts_1 >0,
           first_team_deficit_pts_2 < first_team_deficit_pts_1,
           first_team_deficit_pts_2 <= 0)



#biggest_gd_chases = all_gd_chases %>%
#  ungroup() %>%
#  #look at max chase by team pair per season
#  #i.e. between team pair, with minimum chase time (i.e. latest week 1, earliest week 2)
#    group_by(first_team, second_team) %>%
#    filter(chase == max(chase)) %>%
#    filter(gameweek_2 == min(gameweek_2)) %>%
#    filter(gameweek_1 == max(gameweek_1)) %>%
#  #now look at biggest chase between any pair in the season
#    ungroup() %>%
#    arrange(desc(chase)) %>%
#    filter(chase == max(chase)) 

# Fill the summary summary dataframes
all_gd_pts_evolution_all_seasons = all_gd_pts_evolution_all_seasons %>%
  bind_rows(all_gd_pts_evolution)

all_gd_chases_all_seasons = all_gd_chases_all_seasons %>%
  bind_rows(all_gd_chases)

all_pts_chases_all_seasons = all_pts_chases_all_seasons %>%
  bind_rows(all_pts_chases)

}

```


## GAMEWEEK VERSION OF THE ABOVE - CHASES BY GAMEWEEK, NOT DATE

```{r}

# This is where the data is from: http://www.football-data.co.uk/englandm.php

# Create blank summary dataframe to fill with the loop
all_gd_pts_evolution_all_seasons_gw = data.frame()
all_gd_chases_all_seasons_gw = data.frame()
all_pts_chases_all_seasons_gw = data.frame()
all_gd_and_pts_chases_all_seasons_gw = data.frame()
biggest_gd_chases_by_season_gw = data.frame()
first_year = 1994

## LOTS OF MANUAL CHANGES NEEDED TO MOVE FROM GAMEWEEK TO DATE BASED CALCS
#  AND VICE VERSA


# Loop to calculate running goal difference and points
# And then caclulate chases
while (first_year < 2021) {
  
  # Define year and create URL for that year
  first_year = first_year+1
  second_year = first_year+1
  first_year_trimmed = str_sub(as.character(first_year), 3, 4)
  second_year_trimmed = str_sub(as.character(second_year), 3, 4)
  url = paste("http://www.football-data.co.uk/mmz4281/", 
              first_year_trimmed, second_year_trimmed, "/E0.csv", sep = "")

# Create running points and goal diff totals by team
all_gd_pts_evolution_gw = read_csv(url) %>%
    rename(date = Date) %>%
    filter(date != "") %>%
    select(1:10) %>%
    pivot_longer(cols = c(HomeTeam, AwayTeam), names_to = "venue", values_to = "Team") %>%
    mutate(gd = case_when(venue == "HomeTeam" ~ (FTHG - FTAG),
                                       venue == "AwayTeam" ~ (FTAG - FTHG)),
           pts = case_when(venue == "HomeTeam" & FTHG > FTAG ~ 3,
                              venue == "HomeTeam" & FTHG < FTAG ~ 0,
                              venue == "AwayTeam" & FTAG > FTHG ~ 3,
                              venue == "AwayTeam" & FTAG < FTHG ~ 0,
                              TRUE ~ 1),
           Team = paste(Team, first_year_trimmed, second_year_trimmed,
                        sep = "_")) %>%
    group_by(Team) %>%
    mutate(Gameweek = row_number(),
           gd_running = cumsum(gd),
           pts_running = cumsum(pts)) %>%
    select(date, Gameweek, Team, gd_running, pts_running) %>%

    # For chases by date, rather than gameweek, every club needs a total for...
    # every game date throughout the season, based on games they have played...
    # to that point. This creates full time series of dates for every club...
    # then fills everything downwards
    ungroup() %>%
    mutate(date = lubridate::dmy(date)) %>%
    complete(date, nesting(Team)) %>%
    group_by(Team) %>%
    fill(everything(), .direction = c("down")) %>%
    replace(is.na(.), 0)


# Calculate difference between every team at every Gameweek timestep
all_gd_pts_deficits_gw = all_gd_pts_evolution_gw %>%
    # This section specifies chases will be by Gameweek
    select(-date) %>%
    distinct() %>%
    left_join(x = ., y = ., by = "Gameweek") %>%
  
    filter(Team.x != Team.y) %>%
    mutate(first_team_deficit = gd_running.y - gd_running.x,
           first_team_deficit_pts = pts_running.y - pts_running.x) %>%
    left_join(x = ., y = ., by = c("Team.x" = "Team.x", "Team.y" = "Team.y")) %>%
    rename(first_team = Team.x, second_team = Team.y, 
           first_team_gd_1 = gd_running.x.x, second_team_gd_1 = gd_running.y.x,
           first_team_gd_2 = gd_running.x.y, second_team_gd_2 = gd_running.y.y,
           first_team_deficit_1 = first_team_deficit.x, first_team_deficit_2 = first_team_deficit.y,
           
           first_team_pts_1 = pts_running.x.x, second_team_pts_1 = pts_running.y.x,
           first_team_pts_2 = pts_running.x.y, second_team_pts_2 = pts_running.y.y,
           first_team_deficit_pts_1 = first_team_deficit_pts.x, 
           first_team_deficit_pts_2 = first_team_deficit_pts.y,
           
           # This is specific to Gameweek calculations
           gameweek_1 = Gameweek.x, gameweek_2 = Gameweek.y) %>%
  
    select(first_team, second_team, 
           # This is specific to Gameweek calculations 
           gameweek_1,
           
           first_team_gd_1, second_team_gd_1, first_team_deficit_1,
           first_team_pts_1, second_team_pts_1, first_team_deficit_pts_1,
           # This is specific to Gameweek calculations
           gameweek_2,
           
           first_team_gd_2, second_team_gd_2, first_team_deficit_2,
           first_team_pts_2, second_team_pts_2, first_team_deficit_pts_2) %>%
    # This is specific to Gameweek calculations
    filter(gameweek_2 > gameweek_1) %>%
      
    mutate(swing = first_team_deficit_1 - first_team_deficit_2,
           swing_pts = first_team_deficit_pts_1 - first_team_deficit_pts_2,
           chase = first_team_deficit_1,
           chase_pts = first_team_deficit_pts_1,
           # This is specific to Gameweek calculations
           total_gameweeks = gameweek_2 - gameweek_1)


    
    
# filter so you look at genuine goal diff chases i.e. one team overtakes another
# I.e. there is an inital (positive) gap,
# the second gap is smaller than the first one
# and the second gap is less than/equal to zero
all_gd_chases_gw = all_gd_pts_deficits_gw %>%
    filter(first_team_deficit_1 >0,
           first_team_deficit_2 < first_team_deficit_1,
           first_team_deficit_2 <= 0)

# filter so you look at genuine points chases i.e. one team overtakes another
# I.e. there is an inital (positive) gap,
# the second gap is smaller than the first one
# and the second gap is less than/equal to zero
all_pts_chases_gw = all_gd_pts_deficits_gw %>%
    filter(first_team_deficit_pts_1 >0,
           first_team_deficit_pts_2 < first_team_deficit_pts_1,
           first_team_deficit_pts_2 <= 0)



#biggest_gd_chases = all_gd_chases %>%
#  ungroup() %>%
#  #look at max chase by team pair per season
#  #i.e. between team pair, with minimum chase time (i.e. latest week 1, earliest week 2)
#    group_by(first_team, second_team) %>%
#    filter(chase == max(chase)) %>%
#    filter(gameweek_2 == min(gameweek_2)) %>%
#    filter(gameweek_1 == max(gameweek_1)) %>%
#  #now look at biggest chase between any pair in the season
#    ungroup() %>%
#    arrange(desc(chase)) %>%
#    filter(chase == max(chase)) 

# Fill the summary summary dataframes
all_gd_pts_evolution_all_seasons_gw = all_gd_pts_evolution_all_seasons_gw %>%
  bind_rows(all_gd_pts_evolution_gw)

all_gd_chases_all_seasons_gw = all_gd_chases_all_seasons_gw %>%
  bind_rows(all_gd_chases_gw)

all_pts_chases_all_seasons_gw = all_pts_chases_all_seasons_gw %>%
  bind_rows(all_pts_chases_gw)

}
```


## TIDY THE SUMMARY TABLES FOR GAMEWEEK CHASES

```{r}

#For some reason, the above produces duplicates. 
#Not sure why, but remove them, from gd and pts

#These need to be amended to put date equivalents in too

all_gd_chases_all_seasons_gw = distinct(all_gd_chases_all_seasons_gw) %>%
    ungroup() %>%
    #look at all unique chases 
    #i.e. between a pair of teams, starting on a given week, ending at the earliest possible week
    group_by(first_team, second_team, chase, gameweek_1) %>%
    filter(gameweek_2 == min(gameweek_2)) %>%
    #now do similar thing to find latest start week of the chase
    ungroup() %>%
    group_by(first_team, second_team, chase, gameweek_2) %>%
    filter(gameweek_1 == max(gameweek_1)) %>%
    #now exclude smaller chases that occur within larger ones %>%
    ungroup() %>%
    group_by(first_team, second_team, gameweek_2) %>%
    filter(chase == max(chase)) %>% 
    ungroup() %>%
    group_by(first_team, second_team, gameweek_1) %>%
    filter(chase == max(chase)) %>%
  select(-first_team_pts_1, -second_team_pts_1, -first_team_deficit_pts_1, 
         -first_team_pts_2, -second_team_pts_2, -first_team_deficit_pts_2,
         -swing_pts, -chase_pts)

all_pts_chases_all_seasons_gw = distinct(all_pts_chases_all_seasons_gw) %>%
    ungroup() %>%
    # look at all unique chases 
    # i.e. between a pair of teams, starting on a given week, ending at the earliest possible week
    group_by(first_team, second_team, chase_pts, gameweek_1) %>%
    filter(gameweek_2 == min(gameweek_2)) %>%
    # now do similar thing to find latest start week of the chase
    ungroup() %>%
    group_by(first_team, second_team, chase_pts, gameweek_2) %>%
    filter(gameweek_1 == max(gameweek_1)) %>%
    # now exclude smaller chases that occur within larger ones %>%
    ungroup() %>%
    group_by(first_team, second_team, gameweek_2) %>%
    filter(chase_pts == max(chase_pts)) %>% 
    ungroup() %>%
    group_by(first_team, second_team, gameweek_1) %>%
    filter(chase_pts == max(chase_pts)) %>%
    select(-first_team_gd_1, -second_team_gd_1, -first_team_deficit_1, 
           -first_team_gd_2, -second_team_gd_2, -first_team_deficit_2,
           -swing, -chase)

all_gd_and_pts_chases_all_seasons_gw = all_gd_chases_all_seasons_gw %>%
  inner_join(all_pts_chases_all_seasons_gw)

```

## TIDY THE SUMMARY TABLES FOR DATE CHASES - NEEDS FIXING

```{r}
#For some reason, the above produces duplicates. 
#Not sure why, but remove them, from gd and pts

test = distinct(all_pts_chases_all_seasons) %>%
  ungroup() %>%
  #look at all unique chases 
  #i.e. between a pair of teams, starting on a given week, ending at the earliest possible week
  group_by(first_team, second_team, chase_pts, date_1) %>%
  filter(date_2 == min(date_2)) %>%
  #now do similar thing to find latest start week of the chase
  ungroup() %>%
  group_by(first_team, second_team, chase_pts, date_2) %>%
  filter(date_1 == max(date_1)) %>%
  #now exclude smaller chases that occur within larger ones %>%
  ungroup() %>%
  group_by(first_team, second_team, date_2) %>%
  filter(chase_pts == max(chase_pts)) %>% 
  ungroup() %>%
  group_by(first_team, second_team, date_1) %>%
  filter(chase_pts == max(chase_pts))

test = test %>%
  mutate(pts_per_day = chase_pts/as.numeric(total_days))


test2 = distinct(all_gd_chases_all_seasons) %>%
  ungroup() %>%
  #look at all unique chases 
  #i.e. between a pair of teams, starting on a given week, ending at the earliest possible week
  group_by(first_team, second_team, chase, date_1) %>%
  filter(date_2 == min(date_2)) %>%
  #now do similar thing to find latest start week of the chase
  ungroup() %>%
  group_by(first_team, second_team, chase, date_2) %>%
  filter(date_1 == max(date_1)) %>%
  #now exclude smaller chases that occur within larger ones %>%
  ungroup() %>%
  group_by(first_team, second_team, date_2) %>%
  filter(chase == max(chase)) %>% 
  ungroup() %>%
  group_by(first_team, second_team, date_1) %>%
  filter(chase == max(chase))

test3 = test %>%
  inner_join(test2)

```


## QUICK ATTEMPT TO GGANIMATE THE BIGGEST CHASE

```{r}
all_gd_evolution_all_seasons %>% 
  filter(Team %in% c("Wimbledon_95_96", "Middlesbrough_95_96")) %>%
  mutate(comment = case_when(Gameweek == 19 ~ "After 19 games each, the gap is 24",
                             Gameweek == 38 ~ "By the end of the season, they are level",
                             TRUE ~ "")) %>%
  #Next few lines inspired by this: https://stackoverflow.com/a/53093389
  mutate(show_time = case_when(Gameweek %in% c(19,38) ~ 10,
                               TRUE ~ 1)) %>%
  # uncount is a tidyr function which copies each line 'n' times
  uncount(show_time) %>%
  group_by(Team) %>%
  mutate(reveal_time = row_number()) %>%
  ungroup() %>%
ggplot(aes(Gameweek, gd_running, colour = Team)) + 
  geom_line() +
  geom_point(size = 2) + 
  geom_text(aes(x = Gameweek+1, label = Team), hjust = 0) +
  geom_text(aes(x = Gameweek, y = -5, label = str_wrap(comment, 20), color = "black")) +
  #revealTime rather than gameweek, see above
  transition_reveal(reveal_time) + 
  coord_cartesian(clip = 'off') + 
  labs(title = 'Biggest GD chase in PL History') + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, r = 70, 5.5, 5.5)) +
  theme(legend.position = "none") +
  scale_color_manual(values=c("black",          #for the text
                              "red", "blue"))   #for the two teams

```

## NOW DO ANOTHER INTERESTING CHASE

```{r}
plot_data = all_gd_evolution_all_seasons %>% 
  filter(Team %in% c("Arsenal_20_21", "Aston Villa_20_21")) %>%
  mutate(comment = case_when(Gameweek == 14 ~ "After 14 games each, the gap is 20",
                             Gameweek == 34 ~ "By Gameweek 34, Arsenal have edged ahead",
                             TRUE ~ "")) %>%
  #Next few lines inspired by this: https://stackoverflow.com/a/53093389
  mutate(show_time = case_when(Gameweek %in% c(14,34) ~ 10,
                               TRUE ~ 1)) %>%
  # uncount is a tidyr function which copies each line 'n' times
  uncount(show_time) %>%
  group_by(Team) %>%
  mutate(reveal_time = row_number()) %>%
  ungroup()

ggplot(plot_data, aes(Gameweek, gd_running, colour = Team)) + 
  geom_line() +
  geom_point(size = 2) + 
  geom_text(aes(x = Gameweek+1, label = Team), hjust = 0) +
  #Had a lot of difficulty trying to change the text colour, so keep it black
  geom_text(data = plot_data %>% select(-Team, -gd_running) %>% distinct(),
            aes(x = Gameweek, y = 2, label = str_wrap(comment, 20)),
            inherit.aes = FALSE) +
  coord_cartesian(clip = 'off') + 
  labs(title = 'Biggest GD chase in PL History') + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, r = 70, 5.5, 5.5)) +
  theme(legend.position = "none") +
  scale_color_manual(values=c("#EF0107", "#670E36")) +
  #revealTime rather than gameweek, see above
  transition_reveal(reveal_time) +
  view_follow()

```


## NOW DO ANOTHER INTERESTING CHASE - LIVERPOOL TO UNITED IN 17-18

```{r}
plot_data = all_gd_evolution_all_seasons %>% 
  filter(Team %in% c("Liverpool_17_18", "Man United_17_18")) %>%
  mutate(comment = case_when(Gameweek == 9 ~ "After 9 games each, the gap is 20",
                             Gameweek == 28 ~ "By Gameweek 28, Man U have gained 15, but Liverpool have drawn level",
                             TRUE ~ "")) %>%
  #Next few lines inspired by this: https://stackoverflow.com/a/53093389
  mutate(show_time = case_when(Gameweek %in% c(9,28) ~ 10,
                               TRUE ~ 1)) %>%
  # uncount is a tidyr function which copies each line 'n' times
  uncount(show_time) %>%
  group_by(Team) %>%
  mutate(reveal_time = row_number()) %>%
  ungroup()

ggplot(plot_data, aes(Gameweek, gd_running, colour = Team)) + 
  geom_line() +
  geom_point(size = 2) + 
  geom_text(aes(x = Gameweek+1, label = Team), hjust = 0) +
  #Had a lot of difficulty trying to change the text colour, so keep it black
  geom_text(data = plot_data %>% select(-Team, -gd_running) %>% distinct(),
            aes(x = Gameweek, y = 10, label = str_wrap(comment, 20)),
            inherit.aes = FALSE) +
  coord_cartesian(clip = 'off') + 
  labs(title = 'Liverpool chasing United in 17-18 was highly impressive') + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, r = 70, 5.5, 5.5)) +
  theme(legend.position = "none") +
  scale_color_manual(values=c("#C8102E", "#DA291C")) +
  #revealTime rather than gameweek, see above
  transition_reveal(reveal_time) +
  view_follow()

```


## CALCULATE GD STREAKS (NOT CHASES)

```{r}
#I'm recreating this from an output of the above. Inefficient, better to start from scratch
streak_length  = 5
all_gd_streaks = all_gd_pts_evolution_all_seasons_gw %>%
  select(-date) %>%
  distinct() %>%
  mutate(season = str_sub(Team, -5, -1)) %>%
  group_by(season, Team) %>%
  arrange(Gameweek) %>%
  mutate(gd_gw = gd_running - lag(gd_running),
         pts_gw = pts_running - lag(pts_running),
         gd_streak = RcppRoll::roll_sum(gd_gw, n = streak_length, fill = NA, align = "right"),
         pts_streak = RcppRoll::roll_sum(pts_gw, n = streak_length, fill = NA, align = "right"))
  
  
```


## TESTING

```{r}
test = all_chases_all_seasons %>% 
  mutate(first_team_gain = first_team_gd_2 - first_team_gd_1) %>%
  filter(first_team_gain/chase >= 0.5)
```

